A practical review on linear and nonlinear global approaches to flow instabilities

### 2018-12-07 99_Documentation/ARTICLE_STABFEM/ARTICLE_ASME_Revision.tex
Incorporate corrections sent by David.

\graphicspath{{./}{../../STABLE_CASES/CYLINDER/FIGURES/}}

@@

%%% first author
\author{D. Fabre$^{(a)}$,  V. Citro$^{(b,a)}$,  D. Ferreira Sabino$^{(a)}$, P. Bonnefis$^{(a)}$,  F. Giannetti$^{(b)}$ \& M. Pigou$^{(a)}$ 
    \affiliation{
	$^{(a)}$ Institut de M\'ecanique des fluides de Toulouse (IMFT), University of Toulouse \\
	$^{(b)}$ Dipartimento do Ingegneria (DIIN), Universit\`a} di Salerno. 
}

@@

The stability of fluid flows usually depends on the value of a given parameter.
A bifurcation occurs when a critical value is reached and the original solution becomes linearly unstable, the system then starts evolving towards a new state, either steady or unsteady. 
In the second part of the 19th century, specific analytical and numerical methods have emerged to study these bifurcations and have continuously evolved up to the present days. 
A crucial point, that drove the development in this field, is the availability of significantly increasing and improving computing resources. 
Initially, the linear stability theory focused on fluid flows that are homogeneous in two spatial directions, \textit{e.g.} plane Poiseuille flow \cite{Dreid2004}. In this case, thanks to a Fourier decomposition in the two homogeneous directions, it is possible to reduce the stability problem to a one-dimensional problem, an approach usually referred to as {\em local stability approach}.
On the other hand,  when there are at least two spatial variables, the class of methods suited to solve such problems are generally called {\em global stability approaches}.
A classical example of such behavior in fluid dynamics is the instability occurring in the wake of a circular cylinder. At low Reynolds number (precisely for $Re < 46.7$) the flow is steady and symmetric, but for larger values of Re a global instability arises in the flow field leading to the well-known von K\'arm\'an vortex alley. 
This flow configuration has served as a benchmark in the development of this class of methods.  
If one is only interested in predicting whether a flow is stable, it is sufficient to conduct a {\em linear stability analysis} which is the fundamental brick of global stability approaches. 
Beyond this simple question, in the past two decades, a number of extensions have been developed and popularized.
{\em Adjoint methods} are an important extension \cite{GiannettiLuchini},\cite{Marquet}; they can give insight into the sensitivity of the flow to intrinsic or extrinsic contributions. 
{\em Nonlinear stability approaches} \cite{SippLebedev} \cite{MLugo2014} have also been developed in order to extend the range of applicability of the numerical methods towards large amplitude perturbations.

@@

In the present paper the concepts are introduced and the software is demonstrated for the reference case of the incompressible, two-dimensional flow around a cylinder, but the software is easily customizable to a variety of other situations (compressible, three-dimensional, etc..).

@@

@@

We start from the general problem of a flow field $[{\bf u},p]$ satisfying the incompressible Navier--Stokes equations on a domain $\Omega$,
\begin{eqnarray} \label{NSprimitive}
\partial_t {\bf u} = {\cal NS} ({\bf u},p)
\equiv - {\bf u} \cdot \nabla {\bf u} - \nabla p + \frac{2}{Re}  \nabla \cdot {\mathsf{D}}({\bf u}),  \\
\nabla \cdot {\bf u} = 0,
\end{eqnarray}
with suitable boundary conditions on the frontier $\partial \Omega$ of the domain.
Here $ {\mathsf{D}}({\bf u}) $ is the rate-of-strain tensor defined as
$$
 {\mathsf{D}}({\bf u}) = 1/2
\left( \nabla {\bf u} +  \nabla^T  {\bf u}\right).
$$ 
In the framework of finite element methods, we need to write the equations in weak form.
Prior to this we define a scalar product as follows, for both scalar or vectorial quantities 
$\left< \phi_1, \phi_2 \right> $:
$$
\left< \phi_1, \phi_2 \right> = \int_\Omega \overline{\phi}_1 \cdot \phi_2   \mbox{ d} \Omega.
$$
The weak form of the Navier--Stokes equations is readily defined by introducing test functions 
$[{\bf v},q]$ associated with momentum and continuity equations, and by integrating over the domain\footnote{In the simple presentation given here we have omitted the issue of boundary conditions. Details on how to incorporate boundary conditions in the weak formulation through integration by parts can be found in appendix D.}
\be{NSweak}
\forall [{\bf v},q], \quad \partial_t \left< {\bf v}, {\bf u}\right> = \left< {\bf v} , {\cal NS} ({\bf u},p) \right> + \left< q, \nabla \cdot {\bf u}\right>.
\ee

@@

We look for a steady base-flow $[{\bf u}_b,p_b]$ satisfying the steady Navier--Stokes equations, \textit{i.e.} 
%${\cal NS} ({\bf u}_b,p_b) = 0$.
${\cal NS} ({\bf u}_b,p_b) = 0$.
Suppose that we have a {\em guess}  for the base flow $[{\bf u}_b^g,p_b^g]$  which almost satisfies the equations. 
We look for a better approximation under the form
\be{Newton1}
[{\bf u}_b,p_b]  = [{\bf u}_b^g,p_b^g] + [\delta {\bf u}_b, \delta p_b].
\ee
Injecting \eqref{Newton1} into the weak form \eqref{NSweak} of the Navier--Stokes equations and linearizing leads to  
${\cal NS}  ({\bf u}_b^g,p_b^g) +  {\cal LNS} _{{\bf u}_b^g}(\delta {\bf u}_b,\delta p_b)$, which can also be written in weak form as:
\bes{Newton2}
&\left< {\bf v}, {\cal NS} ({\bf u}_b^g,p_b^g)\right> + \left<q, \nabla \cdot {\bf u}_b^g\right>  
\\
+ &\left< { \bf v}, {\cal LNS} _{{\bf u}_b^g}( \delta {\bf u}_b,\delta p_b) \right> + \left<q, \nabla \cdot \delta{\bf u}_b\right> = 0,
\ees
where ${\cal LNS}$ is the linearised Navier--Stokes operator, defined by its action on a flow field $[{\bf u} , p]$ as follows 
\be{defNSL}
 {\cal LNS}_{{\bf U}}( {\bf u}, p) = - {\cal C}( {\bf U} , {\bf u}) -\nabla p
+\frac{2}{Re} \nabla  \cdot {\mathsf{D}}({\bf u}), %\quad ( \mbox{ with } \nabla \cdot {\bf u } = 0 \, ).
 \ee
and ${\cal C}$ is the convection operator defined by 
\be{defC}
{\cal C}( {\bf U} , {\bf u}) = \left( {\bf U} \cdot \nabla \right) {\bf u} + \left( {\bf u} \cdot \nabla \right)  {\bf U}.
\ee
This problem can now be discretized by projecting upon a basis of Taylor--Hood $(u,v,p) \rightarrow (P2,P2,P1)$ finite elements. Noting $\delta X$ the discretization of $[\delta {\bf u}_b,\delta p_b]$ this eventually leads to a matricial problem of the form $A \cdot \delta X = Y$. The Newton iteration procedure aims at iteratively solving this set of equations up to a defined convergence criterion.
In our implementation, the algorithm for a 2D incompressible flow is written in the Freefem++ solver {\sf Newton\_2D.edp} (an extract of this code featuring the implementation of the Newton loop is given in appendix E). This FreeFem++ solver is wrapped by the Octave/Matlab driver {\sf SF\_BaseFlow.m}.

@@

\paragraph{Adjoint eigenvalue problem, structural sensitivity, and wavemaker characterization}
Developed in the two past decades, the concept of adjoint eigenmodes has now become an unavoidable complement to the linear global stability approach. We here give a short summary of the definition and usefulness of this concept and refer the reader to Luchini \& Bottaro\cite{LucBott2014} for further details.
First of all, the {\em adjoint linearised Navier--Stokes operator} ${\cal LNS}^\dag$ is defined thanks to the following 
property:
\bes{NSLAdj}
\forall ( {\bf u}, p ; {\bf v}, q), & \left< {\cal LNS}^\dag_{\bf U}( {\bf v},q) ,{\bf u}\right> + \left< \nabla \cdot {\bf v},p\right>  \\
=& \left< {\bf v}, {\cal LNS}_{\bf U} ({\bf u},p)\right> + \left< q, \nabla \cdot {\bf u}\right>.
\ees
We can then define the adjoint eigenvalues and eigenmodes as the solutions of the eigenvalue problem 
\be{EigenAdj} 
\forall ( {\bf u}, p), \quad  \lambda^\dag \left< \hat{\bf v}, {\bf u}\right> =
 \left< {\cal LNS}^\dag_{\bf U}( \hat{\bf v},\hat{q}) ,{\bf u}\right> + \left< \nabla \cdot \hat{\bf v},p\right> . 
\ee
It can be shown \cite{SchmidHenningson2001} that the adjoint eigenvalues $\lambda^\dag_k$ are the complex conjugates of the direct eigenvalues $\lambda_k$.

Although the concept of adjoint operator may sound complicated, the resolution of the adjoint problem using finite elements methods is actually extremely easy.
Indeed, as the scalar product used in the definition of the weak formulation \eqref{NSweak} and of the adjoint \eqref{EigenAdj} is the same, the weak formulations of both problems happen to be identical when swapping the test and the unknown functions .
Therefore, the matricial form of the discretized version of \eqref{EigenAdj} is deduced from the one of the direct problem by a simple (Hermitian) transpose of the matrix:
\be{Eigen_Adj_matricial}
\overline{\lambda}^\dag B \hat{X}^\dag = A^T \hat{X}^\dag.
\ee
Adjoint eigenmodes are a powerful tool for investigating problems such as receptivity, transient growth, control and sensitivity (see the reviews of \cite{Jmc2005}, \cite{Ps2007} and \cite{LucBott2014}). The simplest physical interpretation of an adjoint eigenmode is as follows: it corresponds to the initial condition which has maximum projection along the direction of the corresponding eigenmode.
Thus, the adjoint of the most amplified mode corresponds to the optimal perturbation which will maximize the growth of energy in the limit of large time. In effect, one can prove that for $t \rightarrow \infty$  the asymptotic behaviour of a solution with initial condition ${\bf u}_i$  is given as:
$$
{\bf u}(t) \approx 
\frac{ \left< \hat{\bf u}^\dag, {\bf u}_i\right>}{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>} 
e^{\lambda t} \hat{\bf u}.
$$
The choice ${\bf u}_i = \hat{\bf u}^\dag$ is the initial condition of norm unity which maximizes the first factor in this expression.

Investigating the structure of the adjoint eigenmodes gives access to a number of informations about the instability mechanism which are not visible when examining only the direct eigenmodes. First, a simple examination of the dissimilarity between direct and adjoint eigenmodes is an indication of the non-normality of the linear operator, mostly associated to the convection term in open flows \cite{Jmc2005}. Secondly, in a more elaborated way, the product of both fields allows defining the so-called {\em structural sensitivity tensor } defined as 
\be{StructurrTens} 
{\bf{S}}({\bf{x}}) = \frac{ \hat{\bf u}^\dag \otimes \hat{\bf u} }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}}.
\ee

As formalized by \cite{GiannettiLuchini} (following ideas previously proposed by \cite{hill1992theoretical}), this metrics quantifies how an eigenvalue is affected by the introduction of localized feedback of a flow perturbation onto itself. It thereby indicates the spatial regions where closed-loop control using a localized activator will be most efficient. Giannetti \& Luchini (2007) conjectured that regions where such an extrinsic feedback induces the strongest change of the eigenvalue are also the most significant regions considering the intrinsic feedback mechanisms that underpins the genuine eigenmode dynamics. Following this idea, structural sensitivity has thus become a popular way to characherize the  {\em wavemaker region} from where the instability mechanism originates. The structural sensitivity being a tensor, is is convenient to simply represent its norm $S_w$, often simply called 'the wavemaker': 
\be{Structursens} 
S_w({\bf x}) =  {||{\bf{S}}({\bf{x}})||}_\infty \equiv \frac{ ||\hat{\bf u}^\dag({\bf x})|| \,\, ||\hat{\bf u}({\bf x})|| }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}},
\ee 
More recently, \cite{marquet2015endogeneity} reconsidered the question of identifying the active flow regions that drive the instabilities. They argued that such regions might be better identified by a quantity called the {\em endogeneity} and defined as  
\be{Endo} 
S_e({\bf x}) =  \frac{  \hat{\bf u}^\dag({\bf x}) \cdot  {\cal LNS}_{\bf U} \, \hat{\bf u}({\bf x})  }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}}  
= \lambda \frac{ \hat{\bf u}^\dag({\bf x}) \cdot \hat{\bf u}({\bf x}) }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}},
\ee 
This definition has several advantages. In particular, being a complex quantity, it allows distinguishing the flow regions which contribute respectively to the oscillation frequency and the growth rate, hence giving more insight into the intrinsic flow dynamics. This quantity is also potentially useful for mesh design, as discussed in section 2.3.

@@

When a good estimation of the eigenvalue is available, this method is very efficient and converges very rapidly. On the other hand, it can only provide a single eigenvalue. If we want to compute a larger number of eigenvalues, we can revert to a generalized version of iterative methods, called Arnoldi methods\cite{Arnoldi51}. The shift-invert version of the Arnoldi method is in fact the most commonly used method of the current time and is at the basis of both the popular Octave/Matlab function {\tt{eigs}} and the standard eigenvalue solver of FreeFem (\textit{i.e.} ARPACK++). 
%Our implementation in StabFem allows to chose between single eigenvalue computation (power method) and multiple eigenvalue computation (Arnoldi). 
For the case of the 2D incompressible flow, resolution of the eigenvalue problem (including adjoint, sensitivity and endogeneity) is done by the FreeFem++ solver {\sf  Stab2D.edp}; an excerpt of this solver featuring the implementation of the shift-invert algorithm is displayed in appendix E. The latter is wrapped by the Octave/Matlab driver {\sf  SF\_Stability.m} which accepts a number of optional parameters. For instance, the selection between shift-invert and Arnoldi is made according to the parameter \verb|nev|  (\textit{i.e.} number of requested eigenvalues) transmitted to the driver.
(an excerpt of this code featuring the implementation of the Newton loop is given in appendix E).

@@

\begin{figure}
\includegraphics[width=.9 \linewidth]{Cylinder_Lx_baseflow.png}
\includegraphics[width=.9 \linewidth]{Cylinder_Fx_baseflow}
\caption{Recirculation length $L_x$ $(a)$  and nondimensional drag $F_x$ ($b$) of the base flow over a cylinder as function of $Re$.}
\label{fig:LxandDrag}
\end{figure}

@@

As for any numerical method, a crucial aspect for numerical efficiency is the design of the mesh. The finite element method allows using unstructured mesh and hence to locally adapt the refinement. %Following Sipp \& Lebedev, 
The most common procedure is to decompose the domain into several parts with different grid densities; for instance for the wake of a cylinder, we will design a near-wall region with very small  size, a wake region with intermediate mesh size, and an outer region with large mesh size. The drawback of such mesh generation guidelines is that the design relies on an a priori expectation of the regions where gradients will be large.

@@

\section{Illustration for the wake of a cylinder} 
\vspace{.2cm}

\paragraph{Problem description}
Here, we consider the two-dimensional flow of an incompressible fluid of density $\rho$ past a circular cylinder. 
All flow quantities are normalized using the uniform incoming velocity $U_{\infty}$ and the cylinder diameter $D$,  which are the characteristic velocity and length scales used for the definition of Reynolds number $Re= U_{\infty} D / \nu$.
The origin of the Cartesian frame of reference is considered located on the cylinder axis, the x-axis is chosen to be parallel to 
the incoming free-stream velocity while the y-axis with the cross-stream velocity. The dimensions of the computational domain are the following: $-40 \le x/D \le 80$ and $0 \le y/D \le 40$ (boundary conditions and their implementation are detailed in appendix D).
Note that we take advantage of the symmetry properties of the problem to only solve it on the positive $y$ half-domain.

@@

Let us now consider the Octave/Matlab code reported in figure \ref{Listing2}. First we build an initial mesh (line 1), and compute base flow solutions for increasing values of the Reynolds number up to $Re = 60$ (lines 3-9).
Then we perform the mesh adaptation with S strategy, as explained previously (line 13). 
The resulting mesh, depicted in figure \ref{fig:mesh2}, is used for the rest of the computations presented in this paper (except for plotting the structure of the direct eigenmode in figure \ref{fig:Eigenmode}$(a)$ and computing the energy of the nonlinear perturbation displayed in figure \ref{fig:HB_SC_DATA_COMP}$(d)$ which a heavier mesh obtained with adaptation strategy D). 
Appendix A presents additional tests regarding mesh convergence, and demonstrates that results obtained with the resulting mesh are reliable within $0.3\%$ accuracy tolerance for the eigenvalue. It must be emphasized  that the mesh generated through this adaptation process is very light, with only 2048 vertices, which is significantly less than reported in previous work (for instance, in their mesh convergence studies, \cite{SippLebedev} and \cite{MLugo2014} used respectively 190868  and 6731 vertices).

@@

\begin{figure}
\includegraphics[width=.9 \linewidth]{Cylinder_Sigma_Re.png}
\includegraphics[width=.9 \linewidth]{Cylinder_Strouhal_Re.png}
\caption{Growth rate $\sigma$ $(a)$  and Strouhal number $St = \omega/2\pi$ ($b$) as function of Reynolds number}
\label{fig:SigmaOmega}
\end{figure}

\begin{figure}
%\vspace{-.5cm}
\includegraphics[width=.9 \linewidth]{Cylinder_EigenModeRe60_AdaptS.png}%{Cylinder_EigenModeRe60_AdaptD.png}
%\vspace{-.5cm}
%\vspace{-.5cm}

>>>>

\graphicspath{{./}{../../STABLE_CASES/CYLINDER/FIGURES/}{../../DEVELOPMENT_CASES/CompressibleCylinder/Sponge/FIGURES/}}

@@

%%% first author
\author{D. Fabre$^{(a)}$,  V. Citro$^{(b,a)}$,  D. Ferreira Sabino$^{(a)}$, P. Bonnefis$^{(a)}$,  F. Giannetti$^{(b)}$ \& M. Pigou$^{(a)}$ 
    \affiliation{
	$^{(a)}$ Institut de M\'ecanique des fluides de Toulouse (IMFT), University of Toulouse \\
	$^{(b)}$ Dipartimento di Ingegneria (DIIN), Universit\'a} di Salerno. 
}

@@

The stability of fluid flows usually depends on the value of a given parameter.
A bifurcation occurs when a critical value is reached and the original solution becomes linearly unstable, the system then starts evolving towards a new state, either steady or unsteady. 
In the second part of the 19th century, specific analytical and numerical methods have emerged to study these bifurcations and have continuously evolved up to the present days. 
A crucial point, that drove the development in this field, is the availability of significantly increasing and improving computing resources. 
Initially, the linear stability theory focused on fluid flows that are homogeneous in two spatial directions, \textit{e.g.} plane Poiseuille flow \cite{Dreid2004}. In this case, thanks to a Fourier decomposition in the two homogeneous directions, it is possible to reduce the stability problem to a one-dimensional problem, an approach usually referred to as {\em local stability approach}.
On the other hand,  when there are at least two spatial variables, the class of methods suited to solve such problems are generally called {\em global stability approaches}.
A classical example of such behaviour in fluid dynamics is the instability occurring in the wake of a circular cylinder. At low Reynolds number (precisely for $Re < 46.7$) the flow is steady and symmetric, but for larger values of Re a global instability arises in the flow field leading to the well-known von K\'arm\'an vortex alley. 
This flow configuration has served as a benchmark in the development of this class of methods.  
If one is only interested in predicting whether a flow is stable, it is sufficient to conduct a {\em linear stability analysis} which is the fundamental brick of global stability approaches. 
Beyond this simple question, in the past two decades, a number of extensions have been developed and popularized.
{\em Adjoint methods} are an important extension \cite{GiannettiLuchini},\cite{Marquet}; they can give insight into the sensitivity of the flow to intrinsic or extrinsic contributions. 
{\em Nonlinear stability approaches} \cite{SippLebedev} \cite{MLugo2014} have also been developed in order to extend the range of applicability of the numerical methods towards large amplitude perturbations.

@@

In the present paper the concepts are introduced and the software is demonstrated for the reference case of the incompressible, two-dimensional flow around a cylinder, but the software is easily customizable to a variety of other situations (free-surface, three-dimensional, etc..). We show also the application to compressible flows for the same geometrical configuration.

@@

In this part, we expose the main concepts and methods of linear stability analysis for the case of a incompressible flow.
The extension to compressible flow will be considered in section \ref{sec:compressible}.

@@

We start from the general problem of a flow field $[{\bf u},p]$ satisfying the incompressible Navier--Stokes equations on a domain $\Omega$,
\begin{eqnarray} \label{NSprimitive}
\partial_t  [{\bf u},p] = {\cal NS} ( [{\bf u},p] )
\equiv - {\bf u} \cdot \nabla {\bf u} - \nabla p + \frac{2}{Re}  \nabla \cdot {\mathsf{D}}({\bf u}),  \\
\nabla \cdot {\bf u} = 0,
\end{eqnarray}
with suitable boundary conditions on the frontier $\partial \Omega$ of the domain.
Here $ {\mathsf{D}}({\bf u}) $ is the rate-of-strain tensor defined as
$$
 {\mathsf{D}}({\bf u}) = 1/2
\left( \nabla {\bf u} +  \nabla^T  {\bf u}\right).
$$

In the framework of finite element methods, we need to write the equations in weak form.
Prior to this we define a scalar product as follows, for both scalar or vectorial quantities

$\left< \phi_1, \phi_2 \right> $:
$$
\left< \phi_1, \phi_2 \right> = \int_\Omega \overline{\phi}_1 \cdot \phi_2   \mbox{ d} \Omega.
$$
The weak form of the Navier--Stokes equations is readily defined by introducing test functions 
$[{\bf v},q]$ associated with momentum and continuity equations, and by integrating over the domain\footnote{In the simple presentation given here we have omitted the issue of boundary conditions. Details on how to incorporate boundary conditions in the weak formulation through integration by parts can be found in appendix D.}
\be{NSweak}
\forall [{\bf v},q], \quad \partial_t \left< {\bf v}, {\bf u}\right> = \left< {\bf v} , {\cal NS} ({\bf u},p) \right> + \left< q, \nabla \cdot {\bf u}\right>.
\ee

@@

We look for a steady base-flow $[{\bf u}_b,p_b]$ satisfying the steady Navier--Stokes equations, \textit{i.e.} 
%${\cal NS} ({\bf u}_b,p_b) = 0$.
${\cal NS} ({\bf u}_b,p_b) = 0$.
Suppose that we have a {\em guess}  for the base flow $[{\bf u}_b^g,p_b^g]$  which almost satisfies the equations. 
We look for a better approximation under the form
\be{Newton1}
[{\bf u}_b,p_b]  = [{\bf u}_b^g,p_b^g] + [\delta {\bf u}_b, \delta p_b].
\ee

Injecting \eqref{Newton1} into the weak form \eqref{NSweak} of the Navier--Stokes equations and linearizing leads to  
${\cal NS}  ({\bf u}_b^g,p_b^g) +  {\cal LNS} _{{\bf u}_b^g}(\delta {\bf u}_b,\delta p_b) = 0$, which can also be written in weak form as:

\bes{Newton2}
&\left< {\bf v}, {\cal NS} ({\bf u}_b^g,p_b^g)\right> + \left<q, \nabla \cdot {\bf u}_b^g\right>  
\\
+ &\left< { \bf v}, {\cal LNS} _{{\bf u}_b^g}( \delta {\bf u}_b,\delta p_b) \right> + \left<q, \nabla \cdot \delta{\bf u}_b\right> = 0,
\ees
where ${\cal LNS}$ is the linearised Navier--Stokes operator, defined by its action on a flow field $[{\bf u} , p]$ as follows 
\be{defNSL}
 {\cal LNS}_{{\bf U}}( {\bf u}, p) = - {\cal C}( {\bf U} , {\bf u}) -\nabla p
+\frac{2}{Re} \nabla  \cdot {\mathsf{D}}({\bf u}), %\quad ( \mbox{ with } \nabla \cdot {\bf u } = 0 \, ).
 \ee
and ${\cal C}$ is the convection operator defined by 
\be{defC}
{\cal C}( {\bf U} , {\bf u}) = \left( {\bf U} \cdot \nabla \right) {\bf u} + \left( {\bf u} \cdot \nabla \right)  {\bf U}.
\ee

In the framework of finite elements, this problem can now be discretized by projecting upon a basis of Taylor--Hood $(u,v,p) \rightarrow (P2,P2,P1)$ elements. Noting $\delta X$ the discretization of $[\delta {\bf u}_b,\delta p_b]$ this eventually leads to a matricial problem of the form $A \cdot \delta X = Y$. The procedure of Newton iteration is to solve iteratively this set of equations up to convergence.
In our implementation, the algorithm for a 2D incompressible flow is written in the Freefem++ solver {\sf Newton\_2D.edp} (an extract of this code featuring the implementation of the Newton loop is given in appendix E). This FreeFem++ solver is wrapped by the Octave/Matlab driver {\sf SF\_BaseFlow.m}.

@@

\paragraph{Adjoint eigenvalue problem, structural sensitivity and wavemaker characterization}
Developed in the two past decades, the concept of adjoint eigenmodes has now become an unavoidable complement to the linear global stability approach. We here give a short summary of the definition and usefulness of this concept and refer the reader to Luchini \& Bottaro\cite{LucBott2014} for further details.

First of all, the {\em adjoint linearised Navier--Stokes operator} ${\cal LNS}^\dag$ is defined thanks to the following 
property:
\bes{NSLAdj}
\forall ( {\bf u}, p ; {\bf v}, q), & \left< {\cal LNS}^\dag_{\bf U}( {\bf v},q) ,{\bf u}\right> + \left< \nabla \cdot {\bf v},p\right>  \\
=& \left< {\bf v}, {\cal LNS}_{\bf U} ({\bf u},p)\right> + \left< q, \nabla \cdot {\bf u}\right>.
\ees
We can then define the adjoint eigenvalues and eigenmodes as the solution of the eigenvalue problem 
\be{EigenAdj} 
\forall ( {\bf u}, p), \quad  \lambda^\dag \left< \hat{\bf v}, {\bf u}\right> =
 \left< {\cal LNS}^\dag_{\bf U}( \hat{\bf v},\hat{q}) ,{\bf u}\right> + \left< \nabla \cdot \hat{\bf v},p\right> . 
\ee
It can be shown \cite{SchmidHenningson2001} that the adjoint eigenvalues $\lambda^\dag_k$ are the complex conjugates of the direct eigenvalues $\lambda_k$.

Although the concept of adjoint operator may sound complicated, the resolution of the adjoint problem using finite elements methods is actually extremely easy.
Indeed, as the scalar product used in the definition of the weak formulation \eqref{NSweak} and of the adjoint \eqref{EigenAdj} is the same, the weak formulations of both problems happen to be identical when swapping the test and the unknown functions .
Therefore, the matricial form of the discretized version of \eqref{EigenAdj} is deduced from the one of the direct problem by a simple (Hermitian) transpose of the matrix:
\be{Eigen_Adj_matricial}
\overline{\lambda}^\dag B \hat{X}^\dag = A^T \hat{X}^\dag.
\ee
The adjoint field is a powerful tool for investigating problems such as receptivity, transient growth, control and sensitivity (see the reviews of \cite{Jmc2005}, \cite{Ps2007} and \cite{LucBott2014}). The simplest physical interpretation of an adjoint eigenmode is as follows: it corresponds to the initial condition which has maximum projection along the direction of the corresponding eigenmode.
Thus, the adjoint of the most amplified mode corresponds to the optimal perturbation which will maximize the growth of energy in the limit of large time. In effect, one can prove that for $t \rightarrow \infty$  the asymptotic behaviour of a solution with initial condition ${\bf u}_i$  is given as:
$$
{\bf u}(t) \approx 
\frac{ \left< \hat{\bf u}^\dag, {\bf u}_i\right>}{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>} 
e^{\lambda t} \hat{\bf u}.
$$
The choice ${\bf u}_i = \hat{\bf u}^\dag$ is the initial condition of norm unity which maximizes the first factor in this expression.

Investigating the structure of the adjoint eigenmodes gives access to a number of information about the instability mechanism which are not visible when examining only the direct eigenmodes. First, a simple examination of the dissimilarity between direct and adjoint eigenmodes is an indication of the non-normality of the linear operator, mostly associated to the convection term in open flows \cite{Jmc2005}. Secondly, in a more elaborated way, the product of both fields allows defining the so-called {\em structural sensitivity tensor } defined as 
\be{StructurrTens} 
{\bf{S}}({\bf{x}}) = \frac{ \hat{\bf u}^\dag \otimes \hat{\bf u} }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}}.
\ee

As formalized by \cite{GiannettiLuchini} (following ideas previously proposed by \cite{hill1992theoretical}), this metrics quantifies how an eigenvalue is affected by the introduction of localized feedback of a flow perturbation onto itself. It thereby indicates the spatial regions where closed-loop control using a localized activator will be most efficient. Giannetti \& Luchini (2007) conjectured that regions where such an extrinsic feedback induces the strongest change of the eigenvalue are also the most significant regions considering the intrinsic feedback mechanisms that underpins the genuine eigenmode dynamics. Following this idea, structural sensitivity has thus become a popular way to characherize the  {\em wavemaker region} from where the instability mechanism originates. The structural sensitivity being a tensor, is is convenient to simply represent its norm $S_w$, often simply called 'the wavemaker': 
\be{Structursens} 
S_w({\bf x}) =  {||{\bf{S}}({\bf{x}})||}_\infty \equiv \frac{ ||\hat{\bf u}^\dag({\bf x})|| \,\, ||\hat{\bf u}({\bf x})|| }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}},
\ee 
%More recently, \cite{marquet2015endogeneity} reconsidered the question of identifying the active flow regions that drive the instabilities. They argued that such regions might be better identified by a quantity called the {\em endogeneity} and defined as  
%\be{Endo} 
%S_e({\bf x}) =  \frac{  \hat{\bf u}^\dag({\bf x}) \cdot  {\cal LNS}_{\bf U} \, \hat{\bf u}({\bf x})  }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}}  
%= \lambda \frac{ \hat{\bf u}^\dag({\bf x}) \cdot \hat{\bf u}({\bf x}) }{{\left<\hat{\bf u}^\dag, \hat{\bf u}\right>}},
%\ee 
%This definition has several advantages. In particular, being a complex quantity, it allows distinguishing the flow regions which contribute respectively to the oscillation frequency and the growth rate, hence giving more insight into the intrinsic flow dynamics. This quantity is also potentially useful for mesh design, as discussed in section 2.3.

@@

When a good estimation of the eigenvalue is available, this method is very efficient and converges very rapidly. On the other hand, it can only provide a single eigenvalue. If we want to compute a larger number of eigenvalues, we can revert to a generalized version of iterative methods, called Arnoldi methods\cite{Arnoldi51}. The shift-invert version of the Arnoldi method is in fact the most commonly used method of the current time and is at the basis of both the popular Octave/Matlab function {\tt{eigs}} and the standard eigenvalue solver of FreeFem (\textit{i.e.} ARPACK++). 
%Our implementation in StabFem allows to chose between single eigenvalue computation (power method) and multiple eigenvalue computation (Arnoldi).
For the case of the 2D incompressible flow, resolution of the eigenvalue problem is done by the FreeFem++ solver {\sf  Stab2D.edp}; an excerpt of this solver featuring the implementation of the shift-invert algorithm is displayed in appendix E. The latter is wrapped by the Octave/Matlab driver {\sf  SF\_Stability.m} which accepts a number of optional parameters. For instance, the selection between shift-invert and Arnoldi is made according to the parameter \verb|nev|  (\textit{i.e.} number of requested eigenvalues) transmitted to the driver.
(an excerpt of this code featuring the implementation of the Newton loop is given in appendix E).

@@

\begin{figure}
\includegraphics[width=\linewidth]{Cylinder_Lx_baseflow.png}
\includegraphics[width=\linewidth]{Cylinder_Fx_baseflow}
\caption{Recirculation length $L_x$ $(a)$  and nondimensional drag $F_x$ ($b$) of the base flow over a cylinder as function of $Re$.}
\label{fig:LxandDrag}
\end{figure}

@@

As for any numerical method, a crucial aspect for numerical efficiency is the design of the mesh. The finite element method allows using unstructured mesh and hence to have a local adaptation. %Following Sipp \& Lebedev, 
The most common procedure is to decompose the domain into several parts with different grid densities; for instance for the wake of a cylinder, we will design a near-wall region with very small  size, a wake region with intermediate mesh size, and an outer region with large mesh size. The drawback of such mesh generation guidelines is that the design relies on an a priori expectation of the regions where gradients will be large.

@@

\section{Illustration for the wake of a cylinder (incompressible case)}  
\vspace{.2cm}

\paragraph{Problem description}
Here, we consider the two-dimensional flow of an incompressible fluid of density $\rho$ past a circular cylinder. 
All flow quantities are normalized using the uniform incoming velocity $U_{\infty}$ and the cylinder diameter $D$, which are the characteristic velocity and length scales used for the definition of the Reynolds number $Re= U_{\infty} D / \nu$.
The origin of the Cartesian frame of reference is considered located on the cylinder axis, the x-axis is chosen to be parallel to the incoming free-stream velocity while the y-axis with the cross-stream velocity. The dimensions of the computational domain are the following: $-40 \le x/D \le 80$ and $0 \le y/D \le 40$ (boundary conditions and their implementation are detailed in appendix D).
Note that we take advantage of the symmetry properties of the problem to only solve it on the positive $y$ half-domain.

@@

Let us now consider the Octave/Matlab code reported in figure \ref{Listing2}. First we build an initial mesh (line 1), and compute base flow solutions for increasing values of the Reynolds number up to $Re = 60$ (lines 3-9).
Then we perform the mesh adaptation with S strategy, as explained previously (line 13).
The resulting mesh, depicted in figure \ref{fig:mesh2}, is used for the rest of the computations presented in this paper (except for plotting the structure of the direct eigenmode in figure \ref{fig:Eigenmode}$(a)$ and computing the energy of the nonlinear perturbation displayed in figure \ref{fig:HB_SC_DATA_COMP}$(d)$ which adopted a finer mesh obtained with the strategy D). 
Appendix A presents additional tests regarding mesh convergence, and demonstrates that results obtained with the resulting mesh are reliable within $0.3\%$ accuracy tolerance for the eigenvalue. It must be emphasized  that the mesh generated through this adaptation process is very light, with only 2048 vertices, which is significantly less than reported in previous work (for instance, in their mesh convergence studies, \cite{SippLebedev} and \cite{MLugo2014} used respectively 190868  and 6731 vertices).

@@

\begin{figure}
\includegraphics[width=\linewidth]{Cylinder_Sigma_Re.png}
\includegraphics[width=\linewidth]{Cylinder_Strouhal_Re.png}
\caption{Growth rate $\sigma$ $(a)$  and Strouhal number $St = \omega/2\pi$ ($b$) as function of the Reynolds number.}
\label{fig:SigmaOmega}
\end{figure}

\begin{figure}
%\vspace{-.5cm}
\includegraphics[width=\linewidth]{Cylinder_EigenModeRe60_AdaptS.png}%{Cylinder_EigenModeRe60_AdaptD.png}
%\vspace{-.5cm}
%\vspace{-.5cm}

\includegraphics[width=\linewidth]{Cylinder_EigenModeAdjRe60.png}
%\vspace{-.5cm}\vspace{-.5cm}
%\includegraphics[width=\linewidth]{Cylinder_SensitivityRe60.png}
%\vspace{-.5cm}
\caption{Contour plot of the streamwise velocity component: $(a)$ (Direct) Eigenmode; $(b)$ Adjoint mode at $Re= 60$.}
\label{fig:Eigenmode}
\end{figure}
